\part{Contributed Modules}

\chapter{Contributed Modules}

\noindent
Stan is an open-source project and welcomes user contributions.  

In order to reduce maintenance on the main trunk of Stan development
and to allow developer-specified licenses, contributed Stan modules
are not distributed as part of Stan itself.


\section{Contributing a Stan Module}

Developers who have a Stan module to contribute should contact the
Stan developers either through one of the following.
%
\begin{itemize}
\item \code{stan-users} mailing list: 
\\
\url{https://groups.google.com/forum/?fromgroups#!forum/stan-users}
\item Stan e-mail: 
\\
\href{mailto:mc.stanislaw@gmail.com}{\tt mc.stanislaw@gmail.com}
\end{itemize}


\section{GitHub-Hosted Modules}

The \code{stan-dev} organization on GitHub hosts contributed projects
on GitHub.  This is also where the Stan developers will host works in
progress.  The full list of contributed projects on GitHub for
\code{stan-dev} is available at the following location.
%
\begin{quote}
\url{https://github.com/stan-dev}
\end{quote}

Each contributed module on \code{stan-dev}'s GitHub space comes with
its own documentation, indexed by the \code{README.md} file displayed
on GitHub.  Each contributed module has its own licensing the terms of
which are controlled by its developers.  The license for a contributed
package and its dependencies can be found in a top-level directory
\code{licenses/}.

\subsection{Emacs Stan Mode}

\noindent
Emacs Stan mode allows syntax highlighting and automatic indentation
of Stan models in the Emacs text editor.

\begin{quote}
\begin{tabular}{rl}
{\it Repository:} & \url{https://github.com/stan-dev/stan-mode}
\\[4pt]
{\it License:} & GPLv3
\\[4pt]
{\it Authors:} & Jeffrey Arnold, Daniel Lee
\end{tabular}
\end{quote}


