```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(10101010)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
# Posterior Analysis {#analysis.chapter}

Stan uses Markov chain Monte Carlo (MCMC) techniques to generate
samples from the posterior distribution for full Bayesian inference.
Markov chain Monte Carlo (MCMC) methods were developed for situations
in which it is not straightforward to make independent draws
@Metropolis:1953.

In addition to providing point estimates, Stan's optimization algorithm
provides a Laplace approximation from which it is easy to draw
random values.  Stan's variational inference algorithm provides
draws from the variational approximation to the posterior.  Both
of these outputs may be analyzed just as any other MCMC output,
despite the fact that it is atually independent draws.

## Markov Chains

A *Markov chain* is a sequence of random variables $\theta^{(1)},
\theta^{(2)},\ldots$ where each variable is conditionally independent
of all other variables given the value of the previous value.  Thus if
$\theta = \theta^{(1)}, \theta^{(2)},\ldots, \theta^{(N)}$, then

$$
p(\theta) = p(\theta^{(1)}) \prod_{n=2}^N p(\theta^{(n)}|\theta^{(n-1)}).
$$

Stan uses Hamiltonian Monte Carlo to generate a next state in a manner
described in the [Hamiltonian Monte Carlo chapter](#hmc.chapter).

The Markov chains Stan and other MCMC samplers generate are *ergodic*
in the sense required by the Markov chain central limit theorem,
meaning roughly that there is a reasonable chance of reaching
one value of $\theta$ from another.  The Markov chains are also
*stationary*, meaning that the transition probabilities do not change at
different positions in the chain, so that for $n, n' \geq 0$, the
probability function $p(\theta^{(n+1)}|\theta^{(n)})$ is the same as
$p(\theta^{(n'+1)}|\theta^{(n')})$ (following the convention of
overloading random and bound variables and picking out a probability
function by its arguments).

Stationary Markov chains have an *equilibrium distribution* on states
in which each has the same marginal probability function, so that
$p(\theta^{(n)})$ is the same probability function as
$p(\theta^{(n+1)})$.  In Stan, this equilibrium distribution
$p(\theta^{(n)})$ is the target density $p(\theta)$ defined by a Stan
program, which is typically a proper Bayesian posterior density
$p(\theta | y)$ defined on the log scale up to a constant.

Using MCMC methods introduces two difficulties that are not faced by
independent sample Monte Carlo methods.  The first problem is
determining when a randomly initialized Markov chain has converged to
its equilibrium distribution.  The second problem is that the draws
from a Markov chain may be correlated or even anti-correlated, and
thus the central limit theorem's bound on estimation error no longer
applies.  These problems are addressed in the next two sections.

Stan's posterior analysis tools compute a number of summary
statistics, estimates, and diagnostics for Markov chain Monte Carlo
(MCMC) samples.  Stan's estimators and diagnostics are more robust in
the face of non-convergence, antithetical sampling, and long-term
Markov chain correlations than most of the other tools available.  The
algorithms Stan uses to achieve this are described in this chapter.


## Convergence

By definition, a Markov chain generates samples from the target
distribution only after it has converged to equilibrium (i.e.,
equilibrium is defined as being achieved when $p(\theta^{(n)})$ is the
target density).  The following point cannot be expressed strongly
enough:

* In theory, *convergence is only guaranteed asymptotically* as the
number of draws grows without bound.

* In practice, *diagnostics must be applied to monitor convergence*
for the finite number of draws actually available.


## Notation for samples, chains, and draws

To establish basic notation, suppose a target Bayesian posterior
density $p(\theta | y)$ given real-valued vectors of parameters
$\theta$ and real- and discrete-valued data $y$.^[Using vectors simplifies high level exposition at the expense of collapsing structure.] 

An MCMC *sample* consists of a set of a sequence of $M$ Markov chains,
each consisting of an ordered sequence of $N$ *draws* from the
posterior.^[The structure is assumed to be rectangular;  in the
future, this needs to be generalized to ragged samples.]  The sample
thus consists of $M \times N$ draws from the posterior.


### Potential Scale Reduction

One way to monitor whether a chain has converged to the equilibrium
distribution is to compare its behavior to other randomly initialized
chains.  This is the motivation for the @GelmanRubin:1992 potential
scale reduction statistic, $\hat{R}$.  The $\hat{R}$ statistic
measures the ratio of the average variance of samples within each
chain to the variance of the pooled samples across chains; if all
chains are at equilibrium, these will be the same and $\hat{R}$ will
be one.  If the chains have not converged to a common distribution,
the $\hat{R}$ statistic will be greater than one.

Gelman and Rubin's recommendation is that the independent Markov
chains be initialized with diffuse starting values for the parameters
and sampled until all values for $\hat{R}$ are below 1.1.  Stan
allows users to specify initial values for parameters and it is also
able to draw diffuse random initializations automatically satisfying
the declared parameter constraints.

The $\hat{R}$ statistic is defined for a set of $M$ Markov chains,
$\theta_m$, each of which has $N$ samples $\theta^{(n)}_m$.  The
*between-chain variance* estimate is

$$
B
=
\frac{N}{M-1}
\,
\sum_{m=1}^M (\bar{\theta}^{(\bullet)}_{m}
                - \bar{\theta}^{(\bullet)}_{\bullet})^2,
$$

where

$$
\bar{\theta}_m^{(\bullet)}
= \frac{1}{N} \sum_{n = 1}^N \theta_m^{(n)}
$$

and

$$
\bar{\theta}^{(\bullet)}_{\bullet}
= \frac{1}{M} \, \sum_{m=1}^M \bar{\theta}_m^{(\bullet)}.
$$

The *within-chain variance* is averaged over the chains,

$$
W = \frac{1}{M} \, \sum_{m=1}^M s_m^2,
$$

where

$$
s_m^2
=
\frac{1}{N-1}
\, \sum_{n=1}^N (\theta^{(n)}_m - \bar{\theta}^{(\bullet)}_m)^2.
$$

The *variance estimator* is a mixture of the within-chain and
cross-chain sample variances,

$$
\widehat{\mbox{var}}^{+}\!(\theta|y)
= \frac{N-1}{N}\, W \, + \, \frac{1}{N} \, B.
$$

Finally, the *potential scale reduction statistic* is defined by

$$
\hat{R}
\, = \,
\sqrt{\frac{\widehat{\mbox{var}}^{+}\!(\theta|y)}{W}}.
$$


### Split R-hat for Detecting Non-Stationarity

Before Stan calculating the potential-scale-reduction statistic
$\hat{R}$, each chain is split into two halves.  This provides an
additional means to detect non-stationarity in the individual chains.
If one chain involves gradually increasing values and one involves
gradually decreasing values, they have not mixed well, but they can
have $\hat{R}$ values near unity.  In this case, splitting each chain
into two parts leads to $\hat{R}$ values substantially greater than 1
because the first half of each chain has not mixed with the second
half.


### Convergence is Global

A question that often arises is whether it is acceptable to monitor
convergence of only a subset of the parameters or generated
quantities.  The short answer is "no," but this is elaborated
further in this section.

For example, consider the value `lp__`, which is the log posterior
density (up to a constant).^[The `lp__` value also represents the
potential energy in the Hamiltonian system and is rate bounded by the
randomly supplied kinetic energy each iteration, which follows a
Chi-square distribution in the number of parameters.]

It is thus a mistake to declare convergence in any practical sense if
`lp__` has not converged, because different chains are really in
different parts of the space.  Yet measuring convergence for `lp__` is
particularly tricky, as noted below.


#### Asymptotics and transience vs. equilibrium

Markov chain convergence is a global property in the sense that it
does not depend on the choice of function of the parameters that is
monitored.  There is no hard cutoff between pre-convergence
"transience" and post-convergence "equilibrium."  What happens is
that as the number of states in the chain approaches infinity, the
distribution of possible states in the chain approaches the target
distribution and in that limit the expected value of the Monte Carlo
estimator of any integrable function converges to the true
expectation. There is nothing like warmup here, because in the limit,
the effects of initial state are completely washed out.


#### Multivariate convergence of functions

The $\hat{R}$ statistic considers the composition of a Markov chain
and a function, and if the Markov chain has converged then each Markov
chain and function composition will have converged. Multivariate
functions converge when all of their margins have converged by the
Cramer-Wold theorem.

The transformation from unconstrained space to constrained space is
just another function, so does not effect convergence.

Different functions may have different autocorrelations, but if the
Markov chain has equilibrated then all Markov chain plus function
compositions should be consistent with convergence. Formally, any
function that appears inconsistent is of concern and although it would
be unreasonable to test every function, `lp__` and other
measured quantities should at least be consistent.

The obvious difference in `lp__` is that it tends to vary
quickly with position and is consequently susceptible to outliers.


#### Finite numbers of states

The question is what happens for finite numbers of states? If we can
prove a strong geometric ergodicity property (which depends on the
sampler and the target distribution), then one can show that there
exists a finite time after which the chain forgets its initial state
with a large probability. This is both the autocorrelation time and
the warmup time.  But even if you can show it exists and is finite
(which is nigh impossible) you can't compute an actual value
analytically.

So what we do in practice is hope that the finite number of draws is
large enough for the expectations to be reasonably accurate. Removing
warmup iterations improves the accuracy of the expectations but there
is no guarantee that removing any finite number of samples will be
enough.


#### Why inconsistent R-hat?

Firstly, as noted above, for any finite number of draws, there will
always be some residual effect of the initial state, which typically
manifests as some small (or large if the autocorrelation time is huge)
probability of having a large outlier. Functions robust to such
outliers (say, quantiles) will appear more stable and have better
$\hat{R}$. Functions vulnerable to such outliers may show fragility.

Secondly, use of the $\hat{R}$ statistic makes very strong
assumptions. In particular, it assumes that the functions being
considered are Gaussian or it only uses the first two moments and
assumes some kind of independence.  The point is that strong
assumptions are made that do not always hold. In particular, the
distribution for the log posterior density (`lp__`) almost
never looks Gaussian, instead it features long tails that can lead to
large $\hat{R}$ even in the large $N$ limit.  Tweaks to $\hat{R}$,
such as using quantiles in place of raw values, have the flavor of
making the samples of interest more Gaussian and hence the $\hat{R}$
statistic more accurate.


#### Final words on convergence monitoring

"Convergence" is a global property and holds for all integrable
functions at once, but employing the $\hat{R}$ statistic requires
additional assumptions and thus may not work for all functions equally
well.

Note that if you just compare the expectations between chains then we
can rely on the Markov chain asymptotics for Gaussian distributions
and can apply the standard tests.


## Effective Sample Size {#effective-sample-size.section}

The second technical difficulty posed by MCMC methods is that the
samples will typically be autocorrelated (or anticorrelated) within a
chain.  This increases the uncertainty of the estimation of posterior
quantities of interest, such as means, variances, or quantiles; see
@Geyer:2011.

Stan estimates an effective sample size for each parameter, which
plays the role in the Markov chain Monte Carlo central limit theorem
(MCMC CLT) as the number of independent draws plays in the standard
central limit theorem (CLT).

Unlike most packages, the particular calculations used by Stan follow
those for split-$\hat{R}$, which involve both cross-chain (mean) and
within-chain calculations (autocorrelation); see @GelmanEtAl:2013.


### Definition of Effective Sample Size

The amount by which autocorrelation within the chains increases
uncertainty in estimates can be measured by effective sample size (ESS).
Given independent samples, the central limit theorem
bounds uncertainty in estimates based on the number of samples $N$.
Given dependent samples, the number of independent samples is replaced
with the effective sample size $N_{\mathrm{eff}}$, which is
the number of independent samples with the same estimation power as
the $N$ autocorrelated samples.  For example, estimation error is
proportional to $1 / \sqrt{N_{\mathrm{eff}}}$ rather than
$1/\sqrt{N}$.

The effective sample size of a sequence is defined in terms of the
autocorrelations within the sequence at different lags.  The
autocorrelation $\rho_t$ at lag $t \geq 0$ for a chain with joint
probability function $p(\theta)$ with mean $\mu$ and variance
$\sigma^2$ is defined to be

$$
\rho_t
=
\frac{1}{\sigma^2} \, \int_{\Theta} (\theta^{(n)} - \mu)
(\theta^{(n+t)} - \mu) \, p(\theta) \, d\theta.
$$

This is the correlation between the two chains offset by $t$ positions
(i.e., a lag in time-series terminology).  Because we know
$\theta^{(n)}$ and $\theta^{(n+t)}$ have the same marginal
distribution in an MCMC setting, multiplying the two difference terms
and reducing yields

$$
\rho_t
= \frac{1}{\sigma^2}
\, \int_{\Theta}
      \theta^{(n)} \, \theta^{(n+t)} \, p(\theta)
   \, d\theta.
$$

The effective sample size of $N$ samples generated by a process with
autocorrelations $\rho_t$ is defined by
$$
N_{\mathrm{eff}}
\ = \
\frac{N}{\sum_{t = -\infty}^{\infty} \rho_t}
\ = \
\frac{N}{1 + 2 \sum_{t = 1}^{\infty} \rho_t}.
$$

Effective sample size $N_{\mathrm{eff}}$ can be larger than $N$ in
case of antithetic Markov chains, which have negative autocorrelations
on odd lags. The no-U-turn sampling (NUTS) algorithm used in Stan can
produce $N_{\mathrm{eff}}>N$ for parameters which have close to
Gaussian posterior and little dependency on other parameters.


### Estimation of Effective Sample Size

In practice, the probability function in question cannot be tractably
integrated and thus the autocorrelation cannot be calculated, nor the
effective sample size.  Instead, these quantities must be estimated
from the samples themselves.  The rest of this section describes a
autocorrelations and split-$\hat{R}$ based effective sample
size estimator, based on multiple chains. As before, each chain
$\theta_m$ will be assumed to be of length $N$.

Stan carries out the autocorrelation computations for all lags
simultaneously using Eigen's fast Fourier transform (FFT) package with
appropriate padding; see @Geyer:2011 for more detail on using FFT for
autocorrelation calculations.  The autocorrelation estimates
$\hat{\rho}_{t,m}$ at lag $t$ from multiple chains $m \in
(1,\ldots,M)$ are combined with within-sample variance estimate $W$
and multi-chain variance estimate $\widehat{\mbox{var}}^{+}$
introduced in the previous section to compute the combined
autocorrelation at lag $t$ as

$$
\hat{\rho}_t
= 1 - \frac{\displaystyle W
              - \textstyle \frac{1}{M}\sum_{m=1}^M \hat{\rho}_{t,m}}
	    {\widehat{\mbox{var}}^{+}}.
$$

If the chains have not converged, the variance estimator
$\widehat{\mbox{var}}^{+}$ will overestimate variance, leading to an
overestimate of autocorrelation and an underestimate effective sample
size.

Because of the noise in the correlation estimates $\hat{\rho}_t$ as
$t$ increases, a typical truncated sum of $\hat{\rho}_t$ is used.
Negative autocorrelations may occur only on odd lags and by summing
over pairs starting from lag 0, the paired autocorrelation is
guaranteed to be positive, monotone and convex modulo estimator noise
@Geyer:1992, @Geyer:2011.  Stan uses Geyer's initial monotone
sequence criterion. The effective sample size estimator is defined as

$$
\hat{N}_{\mathrm{eff}} = \frac{M \cdot N}{\hat{\tau}},
$$

where

$$
\hat{\tau} = 1 + 2 \sum_{t=1}^{2m+1} \hat{\rho}_t = -1 + 2 \sum_{t'=0}^{m} \hat{P}_{t'},
$$

where $\hat{P}_{t'}=\hat{\rho}_{2t'}+\hat{\rho}_{2t'+1}$. Initial
positive sequence estimators is obtained by choosing the largest $m$
such that $\hat{P}_{t'}>0, \quad t' = 1,\ldots,m$. The initial monotone
sequence is obtained by further reducing $\hat{P}_{t'}$ to the minimum
of the preceding ones so that the estimated sequence is monotone.


### Estimation of MCMC Standard Error

The posterior standard deviation of a parameter $\theta_n$ conditioned
on observed data $y$ is just the standard deviation of the posterior
density $p(\theta_n | y)$.  This is estimated by the standard
deviation of the combined posterior draws across chains,

$$
\hat{\sigma}_n = \mathrm{sd}(\theta^{(1)}_n, \ldots, \theta^{(m)}_n).
$$

The previous section showed how to estimate $N_{\mathrm{eff}}$ for a
parameter $\theta_n$ based on multiple chains of posterior draws.

The mean of the posterior draws of $\theta_n$
$$
\hat{\theta}_n
= \mathrm{mean}(\theta^{(1)}_n, \ldots, \theta^{(m)}_n)
$$

is treated as an estimator of the true posterior mean,

$$
\mathbb{E}[\theta_n \mid y]
\ = \
\int_{-\infty}^{\infty}
    \, \theta \, p(\theta | y)
\, \mathrm{d}\theta_n,
$$

based the observed data $y$.

The standard error for the estimator $\hat{\theta}_n$ is given by the
posterior standard deviation divided by the square root of the
effective sample size.  This standard error is itself estimated as
$\hat{\sigma}_n / \sqrt{N_{\mathrm{eff}}}$.  The smaller the standard
error, the closer the estimate $\hat{\theta}_n$ is expected to be to
the true value.  This is just the MCMC CLT applied to an estimator;
see @Geyer:2011 for more details of the MCMC central limit theorem.


### Thinning Samples

In the typical situation, the autocorrelation, $\rho_t$, decreases as
the lag, $t$, increases.  When this happens, thinning the samples will
reduce the autocorrelation.

For instance, consider generating one thousand posterior draws in one
of the following two ways.

* Generate 1000 draws after convergence and save all of them.

* Generate 10,000 draws after convergence and save every tenth draw.

Even though both produce a sample consisting one thousand draws, the
second approach with thinning can produce a higher effective sample
size.  That's because the autocorrelation $\rho_t$ for the thinned
sequence is equivalent to $\rho_{10t}$ in the unthinned sequence, so
the sum of the autocorrelations will be lower and thus the effective
sample size higher.

Now contrast the second approach above with the unthinned alternative,

* Generate 10,000 draws after convergence and save every draw.

This will have a higher effective sample than the thinned sample
consisting of every tenth drawn.  Therefore, it should be emphasized
that *the only reason to thin a sample is to reduce memory
requirements*.
