```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(10101010)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
# Program Blocks {#blocks.chapter}

A Stan program is organized into a sequence of named blocks, the
bodies of which consist of variable declarations, followed in the case
of some blocks with statements.


## Overview of Stan's Program Blocks

The full set of named program blocks is exemplified in the following
skeletal Stan program.

```stan
functions {
  // ... function declarations and definitions ...
}
data {
  // ... declarations ...
}
transformed data {
   // ... declarations ... statements ...
}
parameters {
   // ... declarations ...
}
transformed parameters {
   // ... declarations ... statements ...
}
model {
   // ... declarations ... statements ...
}
generated quantities {
   // ... declarations ... statements ...
}
```

The function-definition block contains user-defined functions.  The
data block declares the required data for the model.  The transformed
data block allows the definition of constants and transforms of the
data.  The parameters block declares the model's parameters --- the
unconstrained version of the parameters is what's sampled or
optimized.  The transformed parameters block allows variables to be
defined in terms of data and parameters that may be used later and
will be saved.  The model block is where the log probability function
is defined.  The generated quantities block allows derived quantities
based on parameters, data, and optionally (pseudo) random number
generation.


### Optionality and Ordering {-}

All of the blocks are optional.  A consequence of this is that the
empty string is a valid Stan program, although it will trigger a
warning message from the Stan compiler.  The Stan program blocks
that occur must occur in the order presented in the skeletal program
above.  Within each block, both declarations and statements are
optional, subject to the  restriction that the declarations come
before the statements.


### Variable Scope {-}

The variables declared in each block have scope over all subsequent
statements.  Thus a variable declared in the transformed data block
may be used in the model block.  But a variable declared in the
generated quantities block may not be used in any earlier block,
including the model block.  The exception to this rule is that
variables declared in the model block are always local to the model
block and may not be accessed in the generated quantities block; to
make a variable accessible in the model and generated quantities
block, it must be declared as a transformed parameter.

Variables declared as function parameters have scope only within that
function definition's body, and may not be assigned to (they are
constant).


### Function Scope {-}

Functions defined in the function block may be used in any appropriate
block.  Most functions can be used in any block and applied to a
mixture of parameters and data (including constants or program
literals).

Random-number-generating functions are restricted to the generated
quantities block; such functions are suffixed with `_rng`.
Log-probability modifying functions to blocks where the log
probability accumulator is in scope (transformed parameters and
model); such functions are suffixed with `_lp`.

Density functions defined in the program may be used in sampling
statements.


### Automatic Variable Definitions {-}

The variables declared in the `data` and `parameters` block
are treated differently than other variables in that they are
automatically defined by the context in which they are used.  This is
why there are no statements allowed in the data or parameters block.

The variables in the `data` block are read from an external input
source such as a file or a designated R data structure.  The
variables in the `parameters` block are read from the sampler's
current parameter values (either standard HMC or NUTS).  The initial
values may be provided through an external input source, which is also
typically a file or a designated R data structure.  In each case, the
parameters are instantiated to the values for which the model defines
a log probability function.


### Transformed Variables {-}

The `transformed data` and `transformed parameters` block
behave similarly to each other.  Both allow new variables to be
declared and then defined through a sequence of statements.  Because
variables scope over every statement that follows them, transformed
data variables may be defined in terms of the data variables.

Before generating any draws, data variables are read in, then the
transformed data variables are declared and the associated statements
executed to define them.  This means the statements in the transformed
data block are only ever evaluated once.^[If the C++ code is configured for concurrent threads, the data and transformed data blocks can be executed once and reused for multiple chains.]

Transformed parameters work the same way, being defined in terms of
the parameters, transformed data, and data variables.  The difference
is the frequency of evaluation.  Parameters are read in and (inverse)
transformed to constrained representations on their natural scales
once per log probability and gradient evaluation.  This means the
inverse transforms and their log absolute Jacobian determinants are
evaluated once per leapfrog step.  Transformed parameters are then
declared and their defining statements executed once per leapfrog
step.


### Generated Quantities {-}

The generated quantity variables are defined once per sample after all
the leapfrog steps have been completed.  These may be random
quantities, so the block must be rerun even if the Metropolis
adjustment of HMC or NUTS rejects the update proposal.


### Variable Read, Write, and Definition Summary {-}

A table summarizing the point at which variables are read, written, and
defined is given in the [block actions table](#block-actions-table).

**Block Actions Table.**
<a id="block-actions-table"></a>
*The read, write, transform, and evaluate actions and
periodicities listed in the last column correspond to the Stan
program blocks in the first column.  The middle column indicates
whether the block allows statements.  The last row indicates that
parameter initialization requires a read and transform operation
applied once per chain.*

  block   |  statement    |  action / period
:---------|:-------------:|:-------------------
`data` | no | read / chain
`transformed data` | yes | evaluate / chain
`parameters` | no | inv. transform, Jacobian / leapfrog
&nbsp; | &nbsp; | inv. transform, write / sample
`transformed parameters` | yes | evaluate / leapfrog
&nbsp; | &nbsp; | write / sample
`model` | yes | evaluate / leapfrog step
`generated quantities` | yes | eval / sample
&nbsp; | &nbsp; | write / sample
`\slshape (initialization)` | n/a | read, transform / chain

**Variable Declaration Table.**
<a id="where-to-declare-table"></a>
*This table indicates where variables that are not basic
data or parameters should be declared, based on whether it is
defined in terms of parameters, whether it is used in the log
probability function defined in the model block, and whether it is
printed.  The two lines marked with asterisks ($*$) should not be
used as there is no need to print a variable every iteration that
does not depend on the value of any parameters.*

  param depend  |  in target  |  save  |  declare in
:--------------:|:-----------:|:------:|:-----------
 + | + | + | `transformed parameters`
 + | + | - | `model` (local)
 + | - | + | `generated quantities`
 + | - | - | `generated quantities` (local)
 - | + | + | `transformed data` &nbsp; *and* `generated quantities`
 - | + | - | `transformed data`
 - | - | + | `generated quantities`
 - | - | - | `transformed data` (local)

Another way to look at the variables is in terms of their function.
To decide which variable to use, consult the charts in the [variable
declaration table](#where-to-declare-table).  The last line has no
corresponding location, as there is no need to print a variable every
iteration that does not depend on parameters.^[It is possible to print a variable every iteration that does not depend on parameters---just define it (or redefine it if it is transformed data) in the `generated quantities` block.]

The rest of this chapter provides full details on when and how the
variables and statements in each block are executed.


## Statistical Variable Taxonomy

**Statistical Variable Taxonomy Table.**
<a id="variable-kinds-table"></a>
*Variables of the kind indicated in the left column must be declared
in one of the blocks declared in the right column.*

  variable kind    |    declaration block
------------------:|:-----------------------
constants | `data`, `transformed data`
unmodeled data | `data`, `transformed data`
modeled data | `data`, `transformed data`
missing data | `parameters`, `transformed parameters`
modeled parameters | `parameters`, `transformed parameters`
unmodeled parameters | `data`, `transformed data`
derived quantities | `transformed data`, `transformed parameters`, `generated quantities`
loop indices | loop statement

Page 366 of [@GelmanHill:2007] provides a taxonomy of the kinds of
variables used in Bayesian models.  The [table of kinds of
variables](#variable-kinds.figure) contains Gelman and Hill's taxonomy
along with a missing-data kind along with the corresponding locations
of declarations and definitions in Stan.

Constants can be built into a model as literals, data variables, or
as transformed data variables.  If specified as variables, their
definition must be included in data files.  If they are specified as
transformed data variables, they cannot be used to specify the sizes
of elements in the `data` block.

The following program illustrates various variables kinds, listing the
kind of each variable next to its declaration.

```stan
data {
  int<lower=0> N;           // unmodeled data
  real y[N];                // modeled data
  real mu_mu;               // config. unmodeled param
  real<lower=0> sigma_mu;   // config. unmodeled param
}
transformed data {
  real<lower=0> alpha;      // const. unmodeled param
  real<lower=0> beta;       // const. unmodeled param
  alpha = 0.1;
  beta = 0.1;
}
parameters {
  real mu_y;                // modeled param
  real<lower=0> tau_y;      // modeled param
}
transformed parameters {
  real<lower=0> sigma_y;    // derived quantity (param)
  sigma_y = pow(tau_y, -0.5);
}
model {
  tau_y ~ gamma(alpha, beta);
  mu_y ~ normal(mu_mu, sigma_mu);
  for (n in 1:N)
    y[n] ~ normal(mu_y, sigma_y);
}
generated quantities {
  real variance_y;       // derived quantity (transform)
  variance_y = sigma_y * sigma_y;
}
```

In this example, `y[N]` is a modeled data vector.  Although it is
specified in the `data` block, and thus must have a known value
before the program may be run, it is modeled as if it were generated
randomly as described by the model.

The variable `N` is a typical example of unmodeled data.  It is
used to indicate a size that is not part of the model itself.

The other variables declared in the data and transformed data block are
examples of unmodeled parameters, also known as hyperparameters.
Unmodeled parameters are parameters to probability densities that are
not themselves modeled probabilistically.  In Stan, unmodeled
parameters that appear in the `data` block may be specified on a
per-model execution basis as part of the data read.  In the above
model, `mu_mu` and `sigma_mu` are configurable unmodeled
parameters.

Unmodeled parameters that are hard coded in the model must be declared
in the `transformed data` block.  For example, the unmodeled
parameters `alpha` and `beta` are both hard coded to the
value 0.1.  To allow such variables to be configurable based on data
supplied to the program at run time, they must be declared in the
`data` block, like the variables `mu_mu` and
`sigma_mu`.

This program declares two modeled parameters, `mu` and
`tau_y`.  These are the location and precision used in the normal
model of the values in `y`.  The heart of the model will be
sampling the values of these parameters from their posterior
distribution.

The modeled parameter `tau_y` is transformed from a precision to
a scale parameter and assigned to the variable `sigma_y` in the
`transformed parameters` block. Thus the variable `sigma_y`
is considered a derived quantity --- its value is entirely determined
by the values of other variables.

The `generated quantities` block defines a value
`variance_y`, which is defined as a transform of the scale or
deviation parameter `sigma_y`.  It is defined in the generated
quantities block because it is not used in the model.  Making it
a generated quantity allows it to be monitored for convergence (being
a non-linear transform, it will have different autocorrelation and
hence convergence properties than the deviation itself).

In later versions of Stan which have random number generators for
the distributions, the `generated quantities` block will be
usable to generate replicated data for model checking.

Finally, the variable `n` is used as a loop index in the
`model` block.


## Program Block: `data`

The rest of this chapter will lay out the details of each block in
order, starting with the `data` block in this section.


### Variable Reads and Transformations {-}

The `data` block is for the declaration of variables that are
read in as data.  With the current model executable, each Markov chain
of draws will be executed in a different process, and each such
process will read the data exactly once.^[With multiple threads, or even running chains sequentially in a single thread, data could be read only once per set of chains. Stan was designed to be thread safe and future versions will provide a multithreading option for Markov chains.]


Data variables are not transformed in any way.  The format for data
files or data in memory depends on the interface; see the user's
guides and interface documentation for PyStan, RStan, and CmdStan for details.


### Statements {-}

The `data` block does not allow statements.


### Variable Constraint Checking {-}

Each variable's value is validated against its declaration as it is
read.  For example, if a variable `sigma` is declared as
`real<lower=0>`, then trying to assign it a negative value will raise
an error.  As a result, data type errors will be caught as early as
possible.  Similarly, attempts to provide data of the wrong size for a
compound data structure will also raise an error.


## Program Block: `transformed data`

The `transformed data` block is for declaring and defining
variables that do not need to be changed when running the program.


### Variable Reads and Transformations {-}

For the `transformed data` block, variables are all declared in
the variable declarations and defined in the statements.  There is no
reading from external sources and no transformations performed.

Variables declared in the `data` block may be used to declare
transformed variables.


### Statements {-}

The statements in a `transformed data` block are used to define
(provide values for) variables declared in the `transformed data`
block. Assignments are only allowed to variables declared in the
`transformed data` block.

These statements are executed once, in order, right after the data is
read into the data variables.  This means they are executed once per
chain.

Variables declared in the `data` block may be used in statements
in the `transformed data` block.


#### Restriction on Operations in `transformed data` {-}

The statements in the transformed data block are designed to be
executed once and have a deterministic result.  Therefore, log
probability is not accumulated and sampling statements may not be
used.  Random number generating functions are also prohibited.


### Variable Constraint Checking {-}

Any constraints on variables declared in the `transformed data`
block are checked after the statements are executed.  If any defined
variable violates its constraints, Stan will halt with a diagnostic
error message.


## Program Block: `parameters`

The variables declared in the `parameters` program block
correspond directly to the variables being sampled by Stan's samplers
(HMC and NUTS).  From a user's perspective, the parameters in the
program block *are* the parameters being sampled by Stan.

Variables declared as parameters cannot be directly assigned values.
So there is no block of statements in the `parameters` program
block.  Variable quantities derived from parameters may be declared in
the `transformed parameters` or `generated quantities` blocks,
or may be defined as local variables in any statement blocks following
their declaration.

There is a substantial amount of computation involved for parameter
variables in a Stan program at each leapfrog step within the
HMC or NUTS samplers, and a bit more computation along with writes
involved for saving the parameter values corresponding to a sample.


### Constraining Inverse Transform {-}

Stan's two samplers, standard Hamiltonian Monte Carlo (HMC) and the
adaptive No-U-Turn sampler (NUTS), are most easily (and often most
effectively) implemented over a multivariate probability density that
has support on all of $\mathbb{R}^n$.  To do this, the parameters
defined in the `parameters` block must be transformed so they are
unconstrained.

In practice, the samplers keep an unconstrained parameter vector in
memory representing the current state of the sampler.  The model
defined by the compiled Stan program defines an (unnormalized) log
probability function over the unconstrained parameters.  In order to
do this, the log probability function must apply the inverse transform
to the unconstrained parameters to calculate the constrained
parameters defined in Stan's `parameters` program block.  The
log Jacobian of the inverse transform is then added to the accumulated
log probability function.  This then allows the Stan model to be
defined in terms of the constrained parameters.

In some cases, the number of parameters is reduced in the
unconstrained space.  For instance, a $K$-simplex only requires $K-1$
unconstrained parameters, and a $K$-correlation matrix only requires
$\binom{K}{2}$ unconstrained parameters.  This means that the
probability function defined by the compiled Stan program may have
fewer parameters than it would appear from looking at the declarations
in the `parameters` program block.

The probability function on the unconstrained parameters is defined in
such a way that the order of the parameters in the vector corresponds
to the order of the variables defined in the `parameters` program
block.  The details of the specific transformations are provided in
the [variable transforms chapter](#variable-transforms).


### Gradient Calculation {-}

Hamiltonian Monte Carlo requires the gradient of the (unnormalized)
log probability function with respect to the unconstrained parameters
to be evaluated during every leapfrog step.  There may be one leapfrog
step per sample or hundreds, with more being required for models with
complex posterior distribution geometries.

Gradients are calculated behind the scenes using Stan's algorithmic
differentiation library.  The time to compute the gradient does not
depend directly on the number of parameters, only on the number of
subexpressions in the calculation of the log probability.  This
includes the expressions added from the transforms' Jacobians.

The amount of work done by the sampler does depend on the number of
unconstrained parameters, but this is usually dwarfed by the gradient
calculations.


### Writing Draws {-}

In the basic Stan compiled program, there is a file to which the
values of variables are written for each draw.  The constrained
versions of the variables are written in the order they are
defined in the `parameters` block.  In order to do this, the
transformed parameter, model, and generated quantities statements must
also be executed.


## Program Block: `transformed parameters`

The `transformed parameters` program block consists of optional
variable declarations followed by statements.  After the statements
are executed, the constraints on the transformed parameters are
validated.  Any variable declared as a transformed parameter is part
of the output produced for draws.

Any variable that is defined wholly in terms of data or transformed
data should be declared and defined in the transformed data block.
Defining such quantities in the transformed parameters block is legal,
but much less efficient than defining them as transformed data.


### Constraints are for Error Checking {-}

Like the constraints on data, the constraints on transformed
parameters is meant to catch programming errors as well as convey
programmer intent.  They are not automatically transformed in such a
way as to be satisfied.  What will happen if a transformed parameter
does not match its constraint is that the current parameter values
will be rejected.  This can cause Stan's algorithms to hang or
to devolve to random walks.  It is not intended to be a way to enforce
ad hoc constraints in Stan programs.   See the [section on reject statements](#reject-statements)
for further discussion of the behavior of reject statements.


## Program Block: `model`

The `model` program block consists of optional variable
declarations followed by statements.  The variables in the model block
are local variables and are not written as part of the output.

Local variables may not be defined with constraints because there is
no well-defined way to have them be both flexible and easy to
validate.

The statements in the model block typically define the model.  This is
the block in which probability (sampling notation) statements are
allowed.  These are typically used when programming in the BUGS idiom
to define the probability model.


## Program Block: `generated quantities`

The `generated quantities` program block is rather different than
the other blocks.  Nothing in the generated quantities block affects
the sampled parameter values.  The block is executed only after a
sample has been generated.

Among the applications of posterior inference that can be coded in the
generated quantities block are

* forward sampling to generate simulated data for model testing,
* generating predictions for new data,
* calculating posterior event probabilities, including multiple comparisons,
  sign tests, etc.,
* calculating posterior expectations,
* transforming parameters for reporting,
* applying full Bayesian decision theory,
* calculating log likelihoods, deviances, etc. for model comparison.

Parameter estimates, predictions, statistics, and event probabilities
calculated directly using plug-in estimates.  Stan automatically
provides full Bayesian inference by producing draws from the
posterior distribution of any calculated event probabilities,
predictions, or statistics.

Within the generated quantities block, the values of all other variables
declared in earlier program blocks (other than local variables) are
available for use in the generated quantities block.

It is more efficient to define a variable in the generated quantities
block instead of the transformed parameters block.  Therefore, if a
quantity does not play a role in the model, it should be defined in
the generated quantities block.

After the generated quantities statements are executed, the constraints
on the declared generated quantity variables are validated.

All variables declared as generated quantities are printed as part of
the output.
