```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(10101010)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
# Deprecated Features {#deprecated-features.appendix}

This appendix lists currently deprecated functionality along with how
to replace it.  These deprecated features are likely to be removed in
the next major release.  


## Assignment with `<-`

*Deprecated*: The deprecated syntax uses the operator `<-` for
  assignment, e.g.,

```
a <- b;
```

*Replacement*: The new syntax uses the operator `=` for assignment,
  e.g.,
  
```
a = b;
```


## `increment_log_prob` Statement


*Deprecated*: The deprecated syntax for incrementing the log density
  accumulator by `u` is

```
increment_log_prob(u);
```

If `u` is an expression of real type, the underlying log density
accumulator is incremented by `u`; if `u` is a container, the
underlying log density is incremented with each element.

*Replacement*: Replace the above statement with

```
target += u;
```


## `lp__` Variable

*Deprecated*: The variable `lp__` is available wherever log density
increment statements are allowed (`target~+=` and `~` shorthand
statements).

*Replacement*: General manipulation of `lp__` is not allowed, but

```
lp__ <- lp__ + e;
```

can be replaced with

```
target += e;
```

The value of `lp__` is available through the no-argument function
`target()`.


## `get_lp()` Function


*Deprecated*: The no-argument function `get_lp()` is deprecated.

*Replacement*: Use the no-argument function `target()` instead.


## `_log` Density and Mass Functions


*Deprecated*: The probability function for the distribution `foo` will
be applied to an outcome variable `y` and sequence of zero or more
parameters `...` to produce the expression `foo_log(y, ...)`.

*Replacement*: If `y` can be a real value (including vectors
or matrices), replace

```
foo_log(y, ...)
```

with the log probability density function notation

```
foo_lpdf(y | ...).
```

If `y` must be an integer (including arrays), instead replace

```
foo_log(y, ...
```

with the log probability mass function

```
foo_lpmf(y | ...).
```


## `cdf_log` and `ccdf_log` Cumulative Distribution Functions


*Deprecated*: The log cumulative distribution and complementary
cumulative distribution functions for a distribution `foo` are
currently written as `foo_cdf_log` and `foo_ccdf_log`.

*Replacement*: Replace `foo_cdf_log(y, ...)` with `foo_lcdf(y | ...)`.
Replace `foo_ccdf_log(y, ...)` with `foo_lccdf(y | ...)`.


## `multiply_log` and `binomial_coefficient_log` Functions

*Deprecated*: Currently two non-conforming functions ending in suffix
  `_log`.

*Replacement*: Replace `multiply_log(...)` with `lmultiply(...)`.
 Replace `binomial_coefficient_log(...)` with `lchoose(...)`.


## User-Defined Function with `_log` Suffix

*Deprecated*: A user-defined function ending in `_log` can
  be used in sampling statements, with

```
y ~ foo(...);
```

having the same effect as

```
target += foo_log(y, ...);
```

*Replacement*: Replace the `_log` suffix with `_lpdf` for density
functions or `_lpmf` for mass functions in the user-defined function.


## `lkj_cov` Distribution

*Deprecated*:The distribution `lkj_cov` is deprecated.

*Replacement*: Replace `lkj_cov_log(...)` with an `lkj_corr`
  distribution on the correlation matrix and independent lognormal
  distributions on the scales.  That is, replace

```
cov_matrix[K] Sigma;
...
Sigma ~ lkj_cov(mu, tau, eta);
```

with

```
corr_matrix[K] Omega;
vector<lower=0>[K] sigma;
...
Omega ~ lkj_corr(eta);
sigma ~ lognormal(mu, tau);
...
cov_matrix[K] Sigma;
Sigma <- quad_form_diag(Omega, sigma);
```

The variable `Sigma` may be defined as a local variable in the model
block or as a transformed parameter.  An even more efficient transform
would use Cholesky factors rather than full correlation matrix types.


## `if_else` Function

*Deprecated*:The function `if_else` is deprecated.  This function
takes three arguments `a`, `b`, and `c`, where `a` is an `int` value
and `b` and `c` are scalars. It returns `b` if `a` is non-zero and `c`
otherwise.

*Replacement*: Use the conditional operator which allows more
flexibility in the types of `b` and `c` and is much more efficient in
that it only evaluates whichever of `b` or `c` is returned.

```
x = if_else(a,b,c);
```

with

```
x = a ? b : c;
```


## `#` Comments


*Deprecated*: The use of `#` for line-based comments is
  deprecated.  From the first occurrence of `#` onward, the rest
  of the line is ignored.  This happens after includes are resolved
  starting with `#include`.

*Replacement*: Use a pair of forward slashes, `//`, for line
  comments.



