```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(10101010)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
# Diagnostic Mode {#diagnostic-algorithms.chapter}

Stan's diagnostic mode runs a Stan program with data, initializing
parameters either randomly or with user-specified initial values, and
then evaluates the log probability and its gradients. The gradients
computed by the Stan program are compared to values calculated by
finite differences.

Diagnostic mode may be configured with two parameters.

**Diagnostic Mode Configuration Table.**
*The diagnostic model configuration parameters, constraints, and
default values.*

|  parameter    |  description                 |  constraints  |  default |
|:-------------:|:-----------------------------|:-------------:|:--------:|
`epsilon`       | finite difference size       | (0, infty)    |  1e--6   |
`error`         | error threshold for matching | (0, infty)    |  1e--6   |

If the difference between the Stan program's gradient value and that
calculated by finite difference is higher than the specified
threshold, the argument will be flagged.


## Output

Diagnostic mode prints the log posterior density (up to a proportion)
calculated by the Stan program for the specified initial values. For
each parameter, it prints the gradient at the initial parameter values
calculated by Stan's program and by finite differences over Stan's
program for the log probability.


### Unconstrained Scale {-}

The output is for the variable values and their gradients are on the
unconstrained scale, which means each variable is a vector of size
corresponding to the number of unconstrained variables required to
define it. For example, an $N \times N$ correlation matrix, requires
$\binom{N}{2}$ unconstrained parameters. The
transformations from constrained to unconstrained parameters are based
on the constraints in the parameter declarations and described in
the reference manual chapter on transforms.


### Includes Jacobian {-}

The log density includes the Jacobian adjustment implied by the
constraints declared on variables.  The Jacobian adjustment for
constrained parameter transforms will be turned off if optimization is
used in practice, but there is as of yet no way to turn it off in
diagnostic mode.


## Configuration Options

The general configuration options for diagnostics are the same as
those for MCMC. Initial values may be specified, or they may be drawn
at random. Setting the random number generator will only have an
effect if a random initialization is specified.


## Speed Warning and Data Trimming

Due to the application of finite differences, the computation time
grows linearly with the number of parameters. This can be require a
very long time, especially in models with latent parameters that grow
with the data size. It can be helpful to diagnose a model with smaller
data sizes in such cases.
