```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(10101010)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
# Licenses and Dependencies {#licensing.appendix}

Stan and its dependent libraries, are distributed under generous,
freedom-respecting licenses approved by the [Open Source
Initiative](http://opensource.org).

In particular, the licenses for Stan and its dependent libraries have
no ``copyleft'' provisions requiring applications of Stan to be
open source if they are redistributed.

This chapter specifies the licenses for the libraries on which Stan's
math library, language, and algorithms depend.  The last tool
mentioned, Google Test, is only used for testing and is not needed to
run Stan.

## Stan License

Stan is distributed under

* [BSD 3-clause license (BSD New)](http://www.opensource.org/licenses/BSD-3-Clause)}

The copyright holder of each contribution is the developer or his or
her assignee.^[Universities or companies often own the copyright of computer programs developed by their employees.]


## Boost License

Stan uses the Boost library for template metaprograms, traits
programs, the parser, and various numerical libraries for special
functions, probability functions, and random number generators.  Boost
is distributed under the

* [Boost Software License version 1.0](http://www.opensource.org/licenses/BSL-1.0)

The copyright for each Boost package is held by its developers or
their assignees.


## Eigen License

Stan uses the Eigen library for matrix arithmetic and linear algebra.
Eigen is distributed under the 

* [Mozilla Public License, version 2](http://opensource.org/licenses/mpl-2.0)

The copyright of Eigen is owned jointly by its developers or their
assignees.


## SUNDIALS License

Stan uses the SUNDIALS package for solving differential equations.
SUNDIALS is distributed under the

* [BSD 3-clause license (BSD New)](http://www.opensource.org/licenses/BSD-3-Clause)}

The copyright of SUNDIALS is owned by Lawrence Livermore National
Security Lab.


## Google Test License

Stan uses Google Test for unit testing; it is not required to compile
or execute models.  Google Test is distributed under the

* [BSD 3-clause license (BSD New)](http://www.opensource.org/licenses/BSD-3-Clause)}

The copyright of Google Test is owned by Google, Inc.

