```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(123454321)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```
```{r include=FALSE, cache=FALSE}
library(ggplot2)
library(rstan)

set.seed(10101010)

options(digits = 3)

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'left',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```


# Optimization {#optimization-algorithms.chapter}

Stan provides optimization algorithms which find modes of the density
specified by a Stan program. Such modes may be used as parameter
estimates or as the basis of approximations to a Bayesian posterior.

Stan provides three different optimizers, a Newton optimizer, and two
related quasi-Newton algorithms, BFGS and L-BFGS; see
@NocedalWright:2006 for thorough description and analysis of
all of these algorithms. The L-BFGS algorithm is the default
optimizer. Newton's method is the least efficient of the three, but
has the advantage of setting its own stepsize.

## General Configuration

All of the optimizers are iterative and allow the maximum number of
iterations to be specified;  the default maximum number of iterations
is 2000.

All of the optimizers are able to stream intermediate output reporting
on their progress.  Whether or not to save the intermediate iterations
and stream progress is configurable.

## BFGS and L-BFGS Configuration

### Convergence Monitoring {-}

Convergence monitoring in (L-)BFGS is controlled by a number of
tolerance values, any one of which being satisfied causes the
algorithm to terminate with a solution. Any of the convergence tests
can be disabled by setting its corresponding tolerance parameter to
zero.  The tests for convergence are as follows.

#### Parameter Convergence {-}

The parameters $\theta_i$ in iteration $i$ are considered to have
converged with respect to tolerance `tol_param` if

$$
|| \theta_{i} - \theta_{i-1} || < \mathtt{tol\_param}.
$$


#### Density Convergence {-}

The (unnormalized) log density
$\log p(\theta_{i}|y)$ for the parameters $\theta_i$ in iteration $i$
given data $y$ is considered to have converged with
respect to tolerance `tol_obj` if

$$
\left| \log p(\theta_{i}|y) - \log p(\theta_{i-1}|y) \right| <
\mathtt{tol\_obj}.
$$

The log density is considered to have converged to within
relative tolerance `tol_rel_obj` if

$$
\frac{\left| \log p(\theta_{i}|y) - \log p(\theta_{i-1}|y) \right|}{\
  \max\left(\left| \log p(\theta_{i}|y)\right|,\left| \log
      p(\theta_{i-1}|y)\right|,1.0\right)}
 < \mathtt{tol\_rel\_obj} * \epsilon.
$$



#### Gradient Convergence {-}

The gradient is considered to have converged to 0 relative to a
specified tolerance `tol_grad` if

$$
|| g_{i} || < \mathtt{tol\_grad},
$$
where $\nabla_{\theta}$ is the gradient operator with respect to
$\theta$ and $g_{i} = \nabla_{\theta} \log p(\theta | y)$ is the gradient at
iteration $i$ evaluated at $\theta^{(i)}$, the value on the $i$-th posterior iteration.

The gradient is considered to have converged to 0 relative to a
specified relative tolerance
`tol_rel_grad` if

$$
\frac{g_{i}^T \hat{H}_{i}^{-1} g_{i} }{ \max\left(\left|\log
p(\theta_{i}|y)\right|,1.0\right) }
\ < \
\mathtt{tol\_rel\_grad} * \epsilon,
$$

where $\hat{H}_{i}$ is the estimate of the Hessian at iteration $i$,
$|u|$ is the absolute value (L1 norm) of $u$, $||u||$ is the vector
length (L2 norm) of $u$, and $\epsilon \approx 2e-16$ is machine
precision.


### Initial Step Size {-}

The initial step size parameter $\alpha$ for BFGS-style optimizers may
be specified. If the first iteration takes a long time (and requires a
lot of function evaluations) initialize $\alpha$ to be the roughly
equal to the $\alpha$ used in that first iteration. The default value
is intentionally small, 0.001, which is reasonable for many problems
but might be too large or too small depending on the objective
function and initialization. Being too big or too small just means
that the first iteration will take longer (i.e., require more gradient
evaluations) before the line search finds a good step length. It's not
a critical parameter, but for optimizing the same model multiple times
(as you tweak things or with different data), being able to tune
$\alpha$ can save some real time.


### L-BFGS History Size {-}

L-BFGS has a command-line argument which controls the size of the
history it uses to approximate the Hessian. The value should be less than
the dimensionality of the parameter space and, in general, relatively
small values (5--10) are sufficient; the default value is 5.

If L-BFGS performs poorly but BFGS performs well, consider increasing
the history size. Increasing history size will increase the
memory usage, although this is unlikely to be an issue for typical
Stan models.


## General Configuration Options

The general configuration options for optimization are the same as
those for MCMC.


## Writing Models for Optimization

### Constrained vs. Unconstrained Parameters {-}

For constrained optimization problems, for instance, with a standard
deviation parameter $\sigma$ constrained so that $\sigma > 0$, it can
be much more efficient to declare a parameter `sigma` with no
constraints.  This allows the optimizer to easily get close to 0
without having to tend toward $-\infty$ on the $\log \sigma$ scale.

The Jacobian adjustment is not an issue for posterior modes, because
Stan turns off the built-in Jacobian adjustments for optimization.

With unconstrained parameterizations of parameters with constrained
support, it is important to provide a custom initialization that is
within the support.  For example, declaring a vector

```
vector[M] sigma;
```

and using the default random initialization which is
$\mathsf{Uniform}(-2, 2)$ on the unconstrained scale means that there
is only a $2^{-M}$ chance that the initialization will be within
support.

For any given optimization problem, it is probably worthwhile trying
the program both ways, with and without the constraint, to see which
one is more efficient.
